Imperative Programming with Variational Effects

### 2018-09-19 paper/paper.tex
finish refactor
add boxes with types of judgments
ready to submit

\usepackage{amssymb}                                         
\usepackage{amsmath}                                         
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{algorithmic}
\usepackage{url}
\usepackage{tikz}         
\usetikzlibrary{matrix}
\usepackage{listings}
\usepackage{tabu}
\usepackage{mathptmx}
\usepackage{fullpage}
\usepackage{enumerate}
\usepackage{fancyvrb}
\usepackage{mathpartir}
\usepackage{lambda}
\usepackage{cc}
\usepackage{cc-ext}
\usepackage{alltt}
\usepackage{textcomp}
\usepackage{varsem}

@@

\begin{lstlisting}
instance Functor V where
  fmap f (One a) = One (f a)
  fmap f (Chc d l r) = Chc d (fmap f l) (fmap f r)

@@

instance Monad V where
  return = One
  (One v) >>= f = f v
  (Chc d l r) >>= f = Chc d (l >>= f) (r >>= f)
\end{lstlisting}

@@

\begin{samepage}
\begin{algorithmic}
\STATE $y \coloneqq$ getSecret()
\IF{$y$ is true}
\STATE{\textbf{throw} $e$}
\ENDIF
\STATE{$x \coloneqq$ expensiveFn()}
\end{algorithmic}
\end{samepage}

In a non-variational context, the behavior of this program should be clear.
If the variable $y$ evaluates to $\CCkeyw{true}$, then an error is thrown with value
$e$. At this point evaluation should stop, meaning that the variable $x$ is never
assigned in the final statement, avoiding the costly computation of expensiveFn. This effect of halting execution and
returning an error value is the essence of exceptions in non-variational settings.

@@

We also don't want to simply continue evaluation
in every variant regardless of whether or not we have encountered an error.
Suppose that calling \texttt{expensiveFn} normally would evaluate to the variational value $\tagtree{A}{1}{2}$, but at considerable computational cost in both alternatives.
Because we know that the left alternative of $A$ will ultimately evaluate to the thrown exception $e$, we would
like to avoid the cost involved in computing the value $1$ that we will just throw away later, mirroring how throwing an exception short-circuits evaluation in the
non-variational setting.

@@

@@

@@

@@

@@

\begin{figure}[H]
\begin{syntax}
\text{(Store)}\\
\StoreSynt
\text{(Variational integers)}\\
\VIntSynt
\text{(Variational booleans)}\\
\VBoolSynt
\end{syntax}

@@

\begin{figure}[H]
\begin{syntax}
\text{(Store)}\\
S &::=& S(C,x) & \textit{Lookup x in S with context C} \\
& | & S[(C,x) \rightarrow u] & \textit{Update x in S to u with context C} \\
\text{(Variational integers)}\\
\bar{n} &::=& n & \textit{Integer literal} \\
& | & \tagtree{C}{\bar{n}}{\bar{n}} & \textit{Choice} \\
\text{(Variational booleans)}\\
\bar{b} &::=& b & \textit{Boolean literal} \\
& | & \tagtree{C}{\bar{b}}{\bar{b}} & \textit{Choice} \\
\text{(Error lists)} \\
l &::=& [] & \textit{Nil} \\
& | & (C,\bar{n}):l & \textit{Cons} \\
\text{(Error context)} \\
E &::=& [] & \textit{Nil} \\
& | & l : E & \textit{Cons} \\ 
\end{syntax}
\caption{Big-step semantics of Variational Imp statements}
\label{fig:vimpstmt}
\end{figure}

\begin{figure}[H]
\ContinuedFloat
\begin{mathpar}
\BigVSat \and
\BigVUnsat \and
\BigVSkip \and
\BigVAssn \and
\BigVSeq \and
\BigVIf \and
\BigVWhile \and
\BigVThrow \and
\BigVTry \and
\BigVCatch \and
\BigVChc
\end{mathpar}
\caption{Big-step semantics of Variational Imp statements (continued)}
\end{figure}

>>>>

\usepackage{amssymb}                                         
\usepackage{amsmath}                                         
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{algorithmic}
\usepackage{url}
\usepackage{tikz}         
\usetikzlibrary{matrix}
\usepackage{listings}
\usepackage{tabu}
\usepackage{mathptmx}
\usepackage{fullpage}
\usepackage{enumerate}
\usepackage{fancyvrb}
\usepackage{mathpartir}
\usepackage{lambda}
\usepackage{cc}
\usepackage{cc-ext}
\usepackage{alltt}
\usepackage{textcomp}
\usepackage{varsem}
\usepackage{framed}

@@

\begin{minipage}{\linewidth}
\begin{lstlisting}
instance Functor V where
  fmap f (One a) = One (f a)
  fmap f (Chc d l r) = Chc d (fmap f l) (fmap f r)

(<>) :: Functor f => (a -> b) -> f a -> f b  
(<>) = fmap

@@

liftA :: Applicative f => (a -> b) -> f a -> f b
liftA f a = pure f <*> a

liftA2 :: Applicative f => (a -> b -> c) -> f a -> f b -> f c
liftA2 f x = (<*>) (fmap f x)

instance Monad V where
  return = One
  (One v) >>= f = f v
  (Chc d l r) >>= f = Chc d (l >>= f) (r >>= f)
\end{lstlisting}
\end{minipage}

@@

\begin{samepage}
\begin{algorithmic}
\STATE $y \coloneqq$ getSecret()
\IF{$y$ is true}
\STATE{\textbf{throw} $e$}
\ENDIF
\STATE{$x \coloneqq$ expensiveInA()}
\end{algorithmic}
\end{samepage}

In a non-variational context, the behavior of this program should be clear.
If the variable $y$ evaluates to $\CCkeyw{true}$, then an error is thrown with value
$e$. At this point evaluation should stop, meaning that the variable $x$ is never
assigned in the final statement, avoiding the costly computation of expensiveInA. This effect of halting execution and
returning an error value is the essence of exceptions in non-variational settings.

@@

We also don't want to simply continue evaluation
in every variant regardless of whether or not we have encountered an error.
Suppose that calling expensiveInA normally would evaluate to the variational value $\tagtree{A}{1}{2}$, but at considerable computational cost in the left alternative (hence the name).
Because we know that the left alternative of $A$ will ultimately evaluate to the thrown exception $e$, we would
like to avoid the cost involved in computing the value $1$ that we will just throw away later, mirroring how throwing an exception short-circuits evaluation in the
non-variational setting.

@@

\begin{center}
\framebox(200,65){
    \parbox{180\unitlength}{Type of judgments $\Downarrow_A$ and $\Downarrow_B$ \\ $\BigAType$ \\ $\BigBType$}
}
\end{center}

@@

\begin{center}
\framebox(150,50){
    \parbox{125\unitlength}{Type of judgment $\Downarrow_S$ \\ $\BigSType$}
}
\end{center}

@@

\begin{figure}[H]
\begin{syntax}
\text{(Variational optional integers)}\\
\MaybeNumSynt
\text{(Store)}\\
\VStoreSynt
\text{(Variational integers)}\\
\VIntSynt
\text{(Variational booleans)}\\
\VBoolSynt
\end{syntax}

\begin{center}
\framebox(150,50){
    \parbox{125\unitlength}{Type of judgment $\Downarrow_{VA}$ \\ $\BigVAType$}
}
\end{center}

\begin{center}
\framebox(150,50){
    \parbox{125\unitlength}{Type of judgment $\Downarrow_{VB}$ \\ $\BigVBType$}
}
\end{center}

@@

\begin{figure}[H]
\begin{syntax}
\text{(Variational optional integers)}\\
\MaybeNumSynt
\text{(Store)}\\
\VStoreSynt
\text{(Variational integers)}\\
\VIntSynt
\text{(Variational booleans)}\\
\VBoolSynt
\text{(Error lists)} \\
\ErrorListsSynt
\text{(Error context)} \\
\ErrorCtxSynt
\end{syntax}

\begin{center}
\framebox(150,50){
    \parbox{125\unitlength}{Type of judgment $\Downarrow_{V}$ \\ $\BigVType$}
}
\end{center}

\begin{mathpar}
\BigVSat \and
\BigVUnsat
\end{mathpar}

\caption{Big-step semantics of Variational Imp statements}
\label{fig:vimpstmt}
\end{figure}

\begin{center}
\framebox(150,50){
    \parbox{125\unitlength}{Type of judgment $\Downarrow_{VS}$ \\ $\BigVSType$}
}
\end{center}

\begin{mathpar}
\BigVSkip \and
\BigVAssn \and
\BigVSeq \and
\BigVIf \and
\BigVWhile \and
\BigVThrow \and
\BigVTry \and
\BigVCatch \and
\BigVChc
\end{mathpar}
\caption{Big-step semantics of Variational Imp statements (continued)}
\end{figure}

@@

@@

With our definition of both variational execution and configuration for VIMP, we are able to assert that VIMP satisfies the variation-preservation property (Section \ref{sec:props}).
Put simply, the result of evaluating a VIMP program and then configuring it with a set of dimensions should yield the same result as configuring the VIMP program with the
same set of dimensions and then evaluating it using our semantics for plain IMP.
Indeed, we rely on this property throughout our semantics, as highlighted in our explanation above of the behavior of the \textsc{VS-Chc} rule. 
We leave a formal proof that VIMP satisfies this property to future work.
